Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-01-20 paper/paper.tex
Patrick's comments

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices. Such
instances are guaranteed to be satisfiable---\citet{Anton:2009} exploited this property to create
large sets of random satisfiable SAT instances.  However, since this has been the only approach used
to generate random subgraph isomorphism instances, existing benchmark suites contain relatively few
non-trivial unsatisfiable instances, and the satisfiable instances tend to be computationally fairly
easy, with most of the difficulty being in dealing with the size of the model.  This problem cannot
be addressed by using a pattern graph from one of the random suites with the ``wrong'' target graph:
this tends to give either a trivially unsatisfiable instance, or a satisfiable instance. (In
particular, it is \emph{not} the case that a relatively small random graph is unlikely to appear in
a larger random graph.)

@@

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the phase transition for the non-induced variant, for
patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis) and target
(y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in steps of
0.05; for each of these 2601 points, we generate ten random instances. The colour denotes (?? wrong
word?) the proportion of these instances which were found to be satisfiable.  Inside the orange
region, at the bottom right of each plot, every instance is unsatisfiable---here we are trying to
find a dense pattern in a sparse target. In the purple region, at the top left, every instance is
satisfiable---we are looking for a sparse pattern in a dense target (which is easy, since we only
have to preserve adjacency, not non-adjacency). The white band between the regions shows the
location of the phase transition: here, roughly half the instances are satisfiable. (We discuss the
black line below.)

On subsequent rows, we show the average number of search nodes used by the different algorithms. In
general, satisfiable instances are easy, until very close to the phase transition. As we hit the
phase transition and move into the unsatisfiable region, we see a complexity increase. Finally, as
we pass the phase transition and move deeper into the unsatisfiable region, instances become easier
again. This behaviour is largely solver-independent, although VF2 has a larger hard region than
Glasgow or LAD. Thus, although we have moved away from the single control parameter typically
studied, we observe similar behaviour to other \NP-complete problems.

>>>>

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices. Such
instances are guaranteed to be satisfiable---\citet{Anton:2009} exploited this property to create
large sets of random satisfiable boolean satisfiability instances.  However, since this has been the
only approach used to generate random subgraph isomorphism instances, existing benchmark suites
contain relatively few non-trivial unsatisfiable instances, and the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model.  This problem cannot be addressed by using a pattern graph from one of the random suites with
the ``wrong'' target graph: this tends to give either a trivially unsatisfiable instance, or a
satisfiable instance. (In particular, it is \emph{not} the case that a relatively small random graph
is unlikely to appear in a larger random graph.)

@@

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the satisfiability phase transition for the non-induced
variant, for patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis)
and target (y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in
steps of 0.05; for each of these 2601 points, we generate ten random instances. The colour denotes
(?? wrong word?) the proportion of these instances which were found to be satisfiable.  Inside the
orange region, at the bottom right of each plot, every instance is unsatisfiable---here we are
trying to find a dense pattern in a sparse target. In the purple region, at the top left, every
instance is satisfiable---we are looking for a sparse pattern in a dense target (which is easy,
since we only have to preserve adjacency, not non-adjacency). The white band between the regions
shows the location of the phase transition: here, roughly half the instances are satisfiable. (We
discuss the black line below.)

On subsequent rows, we show the average number of search nodes used by the different algorithms. In
general, satisfiable instances are easy, until very close to the phase transition. As we hit the
phase transition and move into the unsatisfiable region, we see a complexity increase. Finally, as
we pass through the phase transition and move deeper into the unsatisfiable region, instances become
easier again. This behaviour is largely solver-independent, although VF2 has a larger hard region
than Glasgow or LAD. Thus, although we have moved away from a single control parameter, we still
observe the easy-hard-easy pattern seen in many other \NP-complete problems.
